Visualizing probability distributions across bivariate cyclic temporal granularities

### 2020-04-24 main.Rmd
english and 'more objective'-type edits from Nick.

abstract: |
 Deconstructing a time index into time granularities can assist in exploration and automated analysis of large temporal data sets. This paper describes several classes of time deconstructions using linear and cyclic time granularities. Linear time granularities respect the linear progression of time such as hours, days, weeks and months with respect to a baseline. Cyclic time granularities can be circular such as hour of the day, quasi-circular such as day of the month, and aperiodic such as public  holidays. The hierarchical structure of granularities creates a nested ordering. Hour of the day and second of the minute are single-order-up. Hour of the week is multiple-order-up, because it passes over day  of the week. Methods are provided for creating all possible granularities for a time index.  A recommendation algorithm provides an indication whether a pair of granularities could be interesting to examine together, called a harmony or clash.

 The time granularities can used to create visualizations of the data to explore for periodicities, associations and anomalies. The granularities can be considered to be categorical variables (ordered and unordered) which induces a grouping of the observations. Assuming a numeric response variable, the resulting graphics are then displays of distributions compared across combinations of categorical variables. A recommendation of appropriate distribution display is provided.

 The methods are implemented in the open source R package gravitas. The functions for creating granularities and exploring the associated time series are consistent with a tidy workflow (@Grolemund2018-po), and the probability distributions can be examined using the range of graphics available in ggplot2 [@Wickham2009pk].
bibliography: [bibliography.bib]
preamble: >
  \setlength {\marginparwidth }{2cm}
  \usepackage{mathtools,amssymb,booktabs,longtable,todonotes,amsthm}
  \def\mod{~\text{mod}~}
output:
  bookdown::pdf_book:
    base_format: rticles::asa_article
    fig_height: 5
    fig_width: 8
    fig_caption: yes
    dev: "pdf"
    keep_tex: yes
---

@@

It is important to be able to navigate through all of these time granularities to have multiple perspectives on the periodicity of the observed data. This idea aligns with the notion of EDA [@Tukey1977-jx] which emphasizes the use of multiple perspectives on data to help formulate hypotheses before proceeding to hypothesis testing. Visualizing probability distributions conditional on one or more granularities is a potentially useful approach for exploration. Analysts are expected to alliteratively explore many possible choices of time granularities for comprehending possible periodicities in the data. But too many choices and a lack of a systematic approach to do so might become overwhelming.

@@

Linear time granularities respect the linear progression of time and are non-repeating such as hours, days, weeks and months. One of the first attempts to characterize these granularities is due to @Bettini1998-ed. However, the definitions and rules defined are inadequate for describing cyclic or repeating granularities. Hence, there is a need to define some new cyclic time granularities, that can be useful in visualizations. Cyclic time granularities can be circular, quasi-circular or aperiodic. Examples of circular granularities are hour of the day and day of the week; an examples of a quasi-circular granularity is day of the month; examples of aperiodic granularities are public holidays and school holidays.

Time deconstructions can also be based on the hierarchical structure of time. For example, hours are nested within days, days within weeks, weeks within months, and so on. Hence, it is possible to construct single-order-up granularities such as second of the minute, or multiple-order-up granularities such as second of the hour. The lubridate package [@Grolemund2011-vm] provides tools to access and manipulate common date-time objects. But most of its accessor functions are limited to single-order-up granularities.

@@

\begin{definition}\label{def:index set}
There is a unique {\bf index set}, $Z=\{z: z \in \mathbb{Z}_{\geq 0}\}$, which map the time instants to a set of positive integers. Often, this is thought of as $t=1, \dots, T$.
\end{definition}

@@

\noindent This implies that the granules in a linear granularity are non-overlapping, continuous and ordering is maintained. The indexing for each granule can also be associated with textual representation, called the label.<!--The label can be mapped to index for input and the index can later be mapped to this label for output. When the label set is exactly the integers, the granularities are called "full-integer" labeled granularity, such as in Definition \ref{def:linear}. Some calendar operations will only work on full-integer labeled granularities, while others can be implemented using more flexible labelings.--> A discrete time model often uses a fixed smallest linear granularity named by @Bettini1998-ed **bottom granularity**.

\autoref{fig:linear-time} illustrates the linear time granularities. Here, "hour" is a linear granularity which is a mapping from index set to hourly time domain. Similarly, "day", "week", "month" and "year" are also linear granularities which maps from index set to subsets of the hourly time domain. For example, day($0$) =  {hour($0$), hour($1$), hour($2$), \dots, hour($23$)}, day($1$) =  {hour($24$), hour($25$), hour($26$), \dots, hour($47$)} and so on. Thus, if we have "hour" running from $\{0, 1, \dots,t\}$, we will have "day" running from $\{0, 1, \dots, \lfloor t/24\rfloor\}$. These linear granularities are ordered with ordering guided by the index set which is a set of integers. Hence, they are uni-directional and non-repeating.

@@

\begin{example}
{\rm Let $day \trianglelefteq week$ and $day \preceq week$ since every granule of $week$ is the union of some set of granules of $day$ and each $day$ is a subset of a $week$. Consider another example where $day \trianglelefteq month$. This relationship however is incomplete without its association to periodicity. Each month is a grouping of the same number of days over years, hence the period of the grouping $(day, month)$ is one year, if leap years are ignored. The period becomes 400 years with the inclusion of leap years and all their exceptions.}
\end{example}

@@

Cyclic granularities represeent cyclical components in time.  Weeks is an example of a cyclical granularity because every 7 days there's another Monday. Cyclic granularities can be constructed by conducting modular arithmetic on two linear granularities. Cycles can be either _regular_, called **circular**, or _irregular_, **quasi-circular**. <!-- with every granule of coarser linear granularity being made up of same (regular) or different (irregular) number of granules of finer linear granularity. For example, a regular mapping exists between  (where 60 minutes always add up to 1 hour), while an irregular mapping exists between  (as the number of days per month ranges from 28 to 31). A grouping operation in Calendar algebra could lead to a regular mapping, whereas a altering-tick operation could result in an irregular mapping. Cyclic time granularities are referred to as **circular** if mappings are regular and **quasi-circular** if the mappings are irregular. Examples of common circular granularities include hour of the day, and day of the week, whereas examples for quasi-circular granularities can be day of the month or week of the month.-->

@@

Cyclic time granularities can be implemented through the tsibble [@wang2019tsibble] data structure. A tsibble consists of an index, key and measured variables. An index is a variable with inherent ordering from past to present and a key is a set of variables that define observational units over time. <!--In a tsibble, each observation (row) is uniquely identified by index and key.--> A time domain, as defined by @Bettini1998-ed, is essentially a mapping of the index set to the index of a tsibble. A linear granularity is a mapping of the index set to subsets of the time domain. For example, if the index of a tsibble is days, then a linear granularity might be weeks, months or years. A bottom granularity is represented by the index of the tsibble.<!--Consistent with the assumption of @Bettini2000-vy, all linear granularities can be generated from the bottom granularity by calendar algebraic operations.-->\todo[inline]{Would it be worth to start this discussion about  data structures earlier, with the linear granularity, and follow it through when appropriate; or put all discussion on  implementation in the data structures section.}

@@

\begin{table}[ht]
\begin{center}
\begin{tabular}{lll}
\hline
circular granularity & expression & period \\
\hline
minute-of-hour                               &
  $C_1 = z \mod 60$                     &
  $P_1 = \phantom{99}60$ \\
minute-of-day                                &
  $C_2 = z \mod 60*24$                  &
  $P_2= `r 60*24`$\\
hour-of-day                                  &
  $C_3 = \lfloor z/60\rfloor\mod 24$    &
  $P_3 = \phantom{99}24$ \\
hour-of-week                                 &
  $C_4 = \lfloor z/60\rfloor\mod 24*7$  &
  $P_4= \phantom{9}`r 24*7`$\\
day-of-week                                  &
  $C_5 = \lfloor z/24*60\rfloor \mod 7$ &
  $P_5= \phantom{999}7$\\
\bottomrule
\end{tabular}
\end{center}
\caption{Examples of circular granularities with bottom granularity minutes. Circular granularity $C_i$ relates two linear granularities one of which groups periodically into the other with regular mapping and period $P_i$. Circular granularities could be expressed using modular arithmetic due to their regular mapping. }
\label{tab:definitions}
\end{table}

@@

<!--Periodic and Quasi-periodic behavior can be defined by a repeating pattern except for a finite number of granules that can be seen as discontinuity points in the granularity in case of quasi-periodic behavior.--> Aperiodic time granularities are the ones which can not be specified as a periodic repetition of a pattern of granules. Most public holidays repeat every year, but there is no reasonably small period within which their behavior remains constant. A classic example can be that of Easter, whose dates repeat only after 5,700,000 years.<!--U.S labor day is the first Monday in September and U.S. Memorial Day is the last day in May. --> In Australia, if a standard public holiday falls on a weekend, a substitute public holiday will sometimes be observed on the first non-weekend day (usually Monday) after the weekend. Examples of aperiodic granularity may also include school holidays or a scheduling event.<!--that might cover the first and third Monday of the month between June and October, except for state holidays.--> All of these are recurring events, but with non-periodic patterns. As such, plausible $P_i$ from \autoref{tab:quasi} could be possibly infinite for aperiodic granularities.

@@

Consider a school semester which always lasts for 18 weeks and 2 days, starting with one week of orientation followed by an in-session period of 6 weeks, a semester-break of 1 week and again an in-session period of 7 weeks and 1 week study break before final exams which continue for 9 days. Let the linear granularities $M_1$ and $M_2$ denote in-session semester period and semester break periods respectively. Both $M_1$, $M_2$ and $M = M_{1}\bigcup M_{2}$ denoting the semester week type are aperiodic with respect to days ($B$) or weeks ($G$). Hence $A_{B, M}$ denoting day of the semester week type would be an aperiodic cyclic granularity. This is because the placement of the semester within an year would vary across years. \autoref{fig:aperiodic-example} b shows the stack display of time with granules representing same categories (semester break/in-session) stacked on top of one another. \autoref{fig:aperiodic-example} (a) representing the linear display of time is useful in manifesting how the categories would be determined with respect to the bottom granularities using Definition \ref{def:aperiodic}. It is interesting to note here that $Q_{H, M}$ denoting semester week of the semester week type would be a quasi-circular granularity since the distribution of semester weeks within a semester is assumed to remain constant over years. Here $Q_{H, M}$ with a period of 18.3 weeks will have irregular mapping since each semester week types consists of different number of semester weeks.

@@

So far we have used example of cyclic granularities from Gregorian calendar as it is the most widely used calendar. But it is far from being the only one. <!--All calendars fall under three types - solar, lunar or lunisolar/solilunar-->The day is the basic unit of time underlying most calendars [@Reingold2001-kf]. Various calendars, however, use different conventions to structure days into larger units: weeks, months, years and cycle of years. The French revolutionary calendar divided each day into 10 "hours", each "hour" into 100 "minutes" and each "minute" into 100 "seconds". Nevertheless, for any calendar a hierarchy table can be defined. For example, in Mayan calendar, one day was referred to as 1 kin and the calendar was structured such that 1 kin = 1 day; 1 uinal = 20 kin; 1 tun  = 18 uinal; 1 katun = 20 tun and 1 baktun = 20 katun. \autoref{tab:tab-mayan} represents the hierarchy table for the Mayan calendar.

@@

Circular single-order-up granularities can be used recursively to obtain multiple order up circular granularity. Since, the operation requires the use of modular arithmetic, it is important that the label mapping of the individual circular single order-up granularity is an identity function, that is, $L(x) = x\quad \forall x$. The label mapping of the resultant multiple-order-up granularity can however be chosen depending on the context.

@@

```{r tab-gregorian}
table_greg <- tibble(`linear (G)` = c("minute", "hour", "day", "month", "year"),
                      `single-order-up cyclic (C)` = c("minute-of-hour","hour-of-day","day-of-month", "month-of-year", 1),
                     `period length/conversion operator (K)` = c(20, 18,"k(day, month)", 20, 1))

kable(table_greg,
      format = "latex",
      booktabs = TRUE,
      caption = "Hierarchy table for Gregorian calendar with both circular and quasi-circular single-order-up granularities.") %>%
  #row_spec(0, bold = TRUE) %>%
   kable_styling()
```

@@

Effective exploration or good visualization require good data structures. It is recognized that there is a crucial influence of linear vs cyclic time characteristics on the expressiveness of visualization and analysis. Moreover, one can use calendars based on application domain that define contextual system of linear and cyclic granularities. A recent tidy data structure to support exploration and modeling of temporal data is tsibble [@wang2019tsibble], where data is structured in a semantic manner with reference to observations and variables, with the time index stated explicitly. Since all cyclic granularities can be expressed in terms of the index set, we consider the data structure in Figure \@ref(fig:data-structure) for exploration of temporal data of this kind. This is a two-dimensional array extending the columns of tsibble by including the cyclic granularities $C_i$ and $v$ denotes the measurement variable. The total number of cyclic granularities would be based on the number of linear granularities considered in the hierarchy table and presence of any aperiodic cyclic granularities. Suppose we have $n$ linear granularities in the hierarchy table, $n(n-1)/2$ circular or quasi-circular cyclic granularities could be constructed. Any attempt to encode all or many cyclic granularities at the same time to develop insights on periodicity might fail or become clumsy. Instead, this big problem is broken down by focusing on two cyclic granularities at a time. Hence data sets of the form <$C_i$, $C_j$, $v$> forms the basis for exploration and analysis of the measured variable.

@@

## Harmonies and Clashes {#sec:synergy}

@@

## Choosing harmonies to show

Even after screening harmonies from the list of all possible combinations of cyclic time granularities that could be plotted together, the options could be too many and it might be useful to rank these harmonies. Ranking could be based on how well a harmony pair displays the variation in distributions of the measured variable within and across each levels of the cyclic granularities. For each facet level, the most interesting display is the one in which the difference in the distribution of the measured variable between the x-axis categories is maximum. And the harmony pair which has more variation between x-axis levels for all facet levels on an average is likely to be a good measure of how important a harmony pair is displaying variation in distributions. Probability distributions are represented through sample percentiles instead of kernel density estimate so that there is minimal dependency on selecting kernel or bandwidth. Moreover, distances between these probability distributions of the measured variable across x-axis levels are computed using Jenson-Shannon distances which is a symmetric divergence obtained from Kullback-Leibler divergence measure.

@@

Restructuring time from linear to cyclic time granularities leads to re-organisation of the data structure, where each level of a cyclic granularity corresponds to multiple values of the measured variable. This is unlike the case of linear granularities where each time point corresponds to an unique value of the measured variable. It is common to see summarization of these multiple values through aggregation or an unique summary statistic like mean or median to have each level of cyclic granularity correspond to an unique value of the measured variable. But that approach  hides the distributions of the measured variable induced by the re-organised data structure. Summarizing the distribution of these multiple observations across cyclic granularities using some summary statistics, each of which can highlight different features of the data or through probability densities could be a potential way to compare and contrast periodicities in the data.<!--Any attempt to encode all or many cyclic granularities together to develop insights on periodicity might become clumsy. Instead, the big problem is broken down by focusing on two cyclic granularities at a time Summarizing properties of the distribution of the measured variable can give valuable indication of features.-->

@@

- Data: <$C_1$, $C_2$, $v$>
- Aesthetic mapping (mapping of variables to elements of the plot): $C_1$ mapped to $x$ position and $v$ to $y$ position
- Statistical transformation (data summarization): Any descriptive or smoothing statistics that summarizes distribution of $v$
- Geometric objects (physical representation of the data): Any geometry displaying distribution, for example, boxplot, letter value, violin, ridge or highest density region plots
- Facet (split plots):  $C_2$

@@

Choice of plots are dictated by the statistical transformations and geometric objects used for the visualization. The basic plot choice for our data structure is the one that can display distributions<!--In Section \ref{sec:measured-variable}, it is discussed that these densities could be estimated.-->. We will discuss few conventional and recent ways to plot distributions using both Kernel density estimates and descriptive statistics. Descriptive statistics based displays include box plots [@Tukey1977-jx] or different variations of it like Notched box plots [@Mcgill1978-hg]. More recent ways are the letter-value box plot [@Hofmann2017-sg] or Quantile plots which display quantiles instead of quartiles in a traditional boxplot. Kernel density based plots for displaying a distribution include violin plots [@Hintze1998-zi], summary plot [@Potter2010-qc], ridge line plots, highest density regions (HDR) box [@Hyndman1996-ft]. Each type of density display has different parameters, that need to be estimated given the data. Each is equipped with some benefits and challenges which should be borne in mind while using them for exploration.

@@

If there are levels of cyclic granularities plotted across x-axis which are not spanned by levels of cyclic granularities plotted across facets or vice versa, we will have empty sets leading to potential ineffective graphs. We hypothesize that the synergy of these cyclic granularities are thus playing a role while deciding if the resulting plot would be a good candidate for exploratory analysis. Harmonies are pairs of granularities that do not contain empty combinations and thus aid exploratory data analysis, and clashes are pairs that do not aid the process of exploratory analysis because they contain empty combinations. As a result, we should avoid plotting clashes. For illustration, \autoref{fig:allFig} (a) shows the distribution of half-hourly electricity consumption through letter value plot across months of the year faceted by quarter-of-year. This plot does not work because every quarter does not correspond to all months of the year, for example, the first quarter would not correspond to the month December of an year.

@@

```{r allFig, out.width="100%", fig.align= 'left',echo=FALSE, eval=TRUE, warning=FALSE,message = FALSE, fig.cap="Distribution of energy consumption displayed through letter value plots. Plot (a) displays consumption across month-of-year faceted by quarter-of-year, (b) across weekday/weekend faceted by quarter-of-year and (c) across quarter-of-year faceted by weekday/weekend. Plots (b) and (c) both show harmonies since each quarter consists of both weekdays and weekends. Conversely, weekend and weekday can occur in every quarter. Plot (b) helps to compare weekends and weekdays for each quarter. It can be seen that for every quarter weekend and weekday consumption are fairly similar except for the second quarter where the letter values below D and E behave differently, whereas, Plot (c) helps to compare quarters within weekdays and weekends. For example, the quartile spread of consumption shrinks/lowers from first to fourth quarter for weekdays, whereas this pattern is not true for weekends. Plot (a) shows a clash since all quarters do not correspond to all months of the year. ", fig.show = 'hold'}

>>>>

abstract: |
 Deconstructing a time index into time granularities can assist in exploration and automated analysis of large temporal data sets. This paper describes several classes of time deconstructions using linear and cyclic time granularities. Linear time granularities respect the linear progression of time such as hours, days, weeks and months with respect to a baseline. Cyclic time granularities can be circular such as hour of the day, quasi-circular such as day of the month, and aperiodic such as public holidays. The hierarchical structure of granularities creates a nested ordering. Hour of the day and second of the minute are single-order-up. Hour of the week is multiple-order-up, because it passes over day of the week. Methods are provided for creating all possible granularities for a time index.  A recommendation algorithm provides an indication whether a pair of granularities can be meaningfully examined together, called a harmony or when they cannot, called a clash.

 The time granularities can be used to create visualizations of the data to explore for periodicities, associations and anomalies. The granularities can be considered to be categorical variables (ordered or unordered) which induces a grouping of the observations. Assuming a numeric response variable, the resulting graphics are then displays of distributions compared across combinations of categorical variables. A recommendation of appropriate distribution display is provided.

 The methods are implemented in the open source R package `gravitas`. The functions for creating granularities and exploring the associated time series are consistent with a tidy workflow (@Grolemund2018-po), and the probability distributions can be examined using the range of graphics available in `ggplot2` [@Wickham2009pk].
bibliography: [bibliography.bib]
preamble: >
  \setlength {\marginparwidth }{2cm}
  \usepackage{mathtools,amssymb,booktabs,longtable,todonotes,amsthm}
  \def\mod{~\text{mod}~}
output:
  bookdown::pdf_book:
    base_format: rticles::asa_article
    fig_height: 5
    fig_width: 8
    fig_caption: yes
    dev: "pdf"
    keep_tex: yes
---

@@

It is important to be able to navigate through all of these time granularities to have multiple perspectives on the periodicity of the observed data. This idea aligns with the notion of EDA [@Tukey1977-jx] which emphasizes the use of multiple perspectives on data to help formulate hypotheses before proceeding to hypothesis testing. Visualizing probability distributions conditional on one or more granularities is an indispensible tool for exploration. Analysts are expected to comprehesively explore the many ways to view and consider temporal data. However, the plethora of choices and the lack of a systematic approach to do so quickly becomes overwhelming.

@@

Linear time granularities respect the linear progression of time and are non-repeating such as hours, days, weeks and months. One of the first attempts to characterize these granularities is due to @Bettini1998-ed. However, the definitions and rules defined are inadequate for describing cyclic or repeating granularities. Hence, there is a need to define some new cyclic time granularities, that can be useful in visualizations. Cyclic time granularities can be circular, quasi-circular or aperiodic. Examples of circular granularities are hour of the day and day of the week; an example of a quasi-circular granularity is day of the month; examples of aperiodic granularities are public holidays and school holidays.

Time deconstructions can also be based on the hierarchical structure of time. For example, hours are nested within days, days within weeks, weeks within months, and so on. Hence, it is possible to construct single-order-up granularities such as second of the minute, or multiple-order-up granularities such as second of the hour. The lubridate package [@Grolemund2011-vm] provides tools to access and manipulate common date-time objects. However, most of its accessor functions are limited to single-order-up granularities.

@@

\begin{definition}\label{def:index set}
There is a unique {\bf index set}, $Z=\{z: z \in \mathbb{Z}_{\geq 0}\}$, which map the time instants to a set of positive integers. Often, this is thought of as $t=0, \dots, T$.
\end{definition}

@@

\noindent This implies that the granules in a linear granularity are non-overlapping, continuous and ordering is maintained. The indexing for each granule can also be associated with textual representation, called the label.<!--The label can be mapped to index for input and the index can later be mapped to this label for output. When the label set is exactly the integers, the granularities are called "full-integer" labeled granularity, such as in Definition \ref{def:linear}. Some calendar operations will only work on full-integer labeled granularities, while others can be implemented using more flexible labelings.--> A discrete time model often uses a fixed smallest linear granularity named by @Bettini1998-ed as the **bottom granularity**.

\autoref{fig:linear-time} illustrates a number of common linear time granularities. Here, "hour" is a linear granularity which is a mapping from index set to hourly time domain. Similarly, "day", "week", "month" and "year" are also linear granularities which maps from index set to subsets of theeir respective time domains. For example, day($0$) =  {hour($0$), hour($1$), hour($2$), \dots, hour($23$)}, day($1$) =  {hour($24$), hour($25$), hour($26$), \dots, hour($47$)} and so on. Thus, if we have "hour" running from $\{0, 1, \dots,t\}$, we will have "day" running from $\{0, 1, \dots, \lfloor t/24\rfloor\}$. These linear granularities are ordered with ordering guided by the index set which is a set of integers. Hence, they are uni-directional and non-repeating.

@@

\begin{example}
{\rm Let $day \trianglelefteq week$ and $day \preceq week$ since every granule of $week$ is the union of some set of granules of $day$ and each $day$ is a subset of a $week$. Consider another example where $day \trianglelefteq month$. This relationship however is incomplete without its association to periodicity. Each month is a grouping of the same number of days over years, hence the period of the grouping $(day, month)$ is one year, if leap years are ignored. This grouping period becomes 4 and 400 years with the inclusion of leap years and leap centries respectively.}
\end{example}

@@

Cyclic granularities represeent cyclical components in time.  Weeks is an example of a cyclical granularity because every 7 days there's another Monday. Cyclic granularities can be constructed by conducting modular arithmetic on two linear granularities. Cycles can be either _regular_, called **circular**, or _irregular_, **quasi-circular**. <!-- with every granule of coarser linear granularity being made up of same (regular) or different (irregular) number of granules of finer linear granularity. For example, a regular mapping exists between  (where 60 minutes always add up to 1 hour), while an irregular mapping exists between  (as the number of days per month ranges from 28 to 31). A grouping operation in Calendar algebra could lead to a regular mapping, whereas a altering-tick operation could result in an irregular mapping. Cyclic time granularities are referred to as **circular** if mappings are regular and **quasi-circular** if the mappings are irregular. Examples of common circular granularities include hour of the day and day of the week, whereas examples for quasi-circular granularities are day of the month and week of the month.-->

@@

Cyclic time granularities can be implemented through the tsibble [@wang2019tsibble] (Time Series taBBLE) data structure. A tsibble consists of an index, key and measured variables. An index is a variable with inherent ordering from past to present and a key is a set of variables that define observational units over time. <!--In a tsibble, each observation (row) is uniquely identified by index and key.--> A time domain, as defined by @Bettini1998-ed, is essentially a mapping of the index set to the index of a tsibble. A linear granularity is a mapping of the index set to subsets of the time domain. For example, if the index of a tsibble is days, then a linear granularity might be weeks, months or years. A bottom granularity is represented by the index of the tsibble.<!--Consistent with the assumption of @Bettini2000-vy, all linear granularities can be generated from the bottom granularity by calendar algebraic operations.-->\todo[inline]{Would it be worth to start this discussion about data structures earlier, with the linear granularity, and follow it through when appropriate; or put all discussion on implementation in the data structures section.}

@@

\begin{table}[ht]
\begin{center}
\begin{tabular}{lll}
\hline
circular granularity & expression & period \\
\hline
minute-of-hour                               &
  $C_1 = z \mod 60$                     &
  $P_1 = \phantom{99}60$ \\
minute-of-day                                &
  $C_2 = z \mod 60*24$                  &
  $P_2= `r 60*24`$\\
hour-of-day                                  &
  $C_3 = \lfloor z/60\rfloor\mod 24$    &
  $P_3 = \phantom{99}24$ \\
hour-of-week                                 &
  $C_4 = \lfloor z/60\rfloor\mod 24*7$  &
  $P_4= \phantom{9}`r 24*7`$\\
day-of-week                                  &
  $C_5 = \lfloor z/24*60\rfloor \mod 7$ &
  $P_5= \phantom{999}7$\\
\bottomrule
\end{tabular}
\end{center}
\caption{Examples of circular granularities with bottom granularity minutes. Circular granularity $C_i$ relates two linear granularities one of which groups periodically into the other with regular mapping and period $P_i$. Circular granularities can be expressed using modular arithmetic due to their regular mapping. }
\label{tab:definitions}
\end{table}

@@

<!--Periodic and Quasi-periodic behavior can be defined by a repeating pattern except for a finite number of granules that can be seen as discontinuity points in the granularity in case of quasi-periodic behavior.--> Aperiodic time granularities are the ones which can not be specified as a periodic repetition of a pattern of granules. Most public holidays repeat every year, but there is no reasonably small period within which their behavior remains constant. A classic example can be that of Easter, whose dates repeat only after 5,700,000 years.<!--U.S labor day is the first Monday in September and U.S. Memorial Day is the last day in May. --> In Australia, if a standard public holiday falls on a weekend, a substitute public holiday will sometimes be observed on the first non-weekend day (usually Monday) after the weekend. Examples of aperiodic granularity may also include school holidays or a scheduled event.<!--that might cover the first and third Monday of the month between June and October, except for state holidays.--> All of these are recurring events, but with non-periodic patterns. As such, plausible $P_i$ from \autoref{tab:quasi} could be possibly infinite for aperiodic granularities.

@@

Consider a school semester which always lasts for 18 weeks and 2 days, starting with one week of orientation followed by an in-session period of 6 weeks, a semester-break of 1 week and again an in-session period of 7 weeks and 1 week study break before final exams which continue for 16 days. Let the linear granularities $M_1$ and $M_2$ denote in-session semester period and semester break periods respectively. Both $M_1$, $M_2$ and $M = M_{1}\bigcup M_{2}$ denoting the semester week type are aperiodic with respect to days ($B$) or weeks ($G$). Hence $A_{B, M}$ denoting day of the semester week type would be an aperiodic cyclic granularity. This is because the placement of the semester within an year would vary across years. \autoref{fig:aperiodic-example} b shows the stack display of time with granules representing same categories (semester break/in-session) stacked on top of one another. \autoref{fig:aperiodic-example} (a) representing the linear display of time is useful in manifesting how the categories would be determined with respect to the bottom granularities using Definition \ref{def:aperiodic}. It is interesting to note here that $Q_{H, M}$ denoting semester week of the semester week type would be a quasi-circular granularity since the distribution of semester weeks within a semester is assumed to remain constant over years. Here $Q_{H, M}$ with a period of 18.3 weeks will have irregular mapping since each semester week types consists of different number of semester weeks.

@@

So far we have used example of cyclic granularities from Gregorian calendar as it is the most widely used calendar. However, it is far from being the only one. <!--All calendars fall under three types - solar, lunar or lunisolar/solilunar-->The day is the basic unit of time underlying most calendars [@Reingold2001-kf]. Various calendars, however, use different conventions to structure days into larger units: weeks, months, years and cycle of years. The French revolutionary calendar divided each day into 10 "hours", each "hour" into 100 "minutes" and each "minute" into 100 "seconds", the duration of which is 0.864 common seconds. Nevertheless, for any calendar a hierarchy table can be defined. For example, in the Mayan calendar, one day is referred to as a kin and the calendar was structured such that 1 kin = 1 day; 1 uinal = 20 kin; 1 tun  = 18 uinal; 1 katun = 20 tun and 1 baktun = 20 katun. \autoref{tab:tab-mayan} represents the hierarchy table for the Mayan calendar.

@@

Circular single-order-up granularities can be used recursively to obtain multiple order up circular granularity. Since, the operation requires the use of modular arithmetic, it is important that the label mapping of the individual circular single order-up granularity is an identity function, that is, $L(x) = x\quad \forall x$. The label mapping of the resultant multiple-order-up granularity can be chosen depending on the context.

@@

```{r tab-gregorian}
table_greg <- tibble(`linear (G)` = c("minute", "hour", "day", "month", "year"),
                      `single-order-up cyclic (C)` = c("minute-of-hour","hour-of-day","day-of-month", "month-of-year", 1),
                     `period length/conversion operator (K)` = c(60, 24, "k(day, month)", 12, 1))

kable(table_greg,
      format = "latex",
      booktabs = TRUE,
      caption = "Hierarchy table for the Gregorian calendar with both circular and quasi-circular single-order-up granularities.") %>%
  #row_spec(0, bold = TRUE) %>%
   kable_styling()
```

@@

Effective exploration and visualization require proper data structures. It is recognized that there is a crucial influence of linear vs cyclic time characteristics on the expressiveness of visualization and analysis. Moreover, one can use calendars based on application domain that define contextual system of linear and cyclic granularities. A recent tidy data structure to support exploration and modeling of temporal data is tsibble [@wang2019tsibble], where data is structured in a semantic manner with reference to observations and variables, with the time index stated explicitly. Since all cyclic granularities can be expressed in terms of the index set, we consider the data structure in Figure \@ref(fig:data-structure) for exploration of temporal data of this kind. This is a two-dimensional array extending the columns of tsibble by including the cyclic granularities $C_i$ and $v$ denotes the measurement variable. The total number of cyclic granularities would be based on the number of linear granularities considered in the hierarchy table and presence of any aperiodic cyclic granularities. Suppose we have $n$ linear granularities in the hierarchy table, $n(n-1)/2$ circular or quasi-circular cyclic granularities could be constructed. Any attempt to encode all or many cyclic granularities at the same time to develop insights on periodicity might fail or otherwise become too numerous for comprehensive human consumption. Instead, this big problem is broken down by focusing on two cyclic granularities at a time. Hence data sets of the form <$C_i$, $C_j$, $v$> forms the basis for exploration and analysis of the measured variable.

@@

## Harmonies and clashes {#sec:synergy}

@@

## Choosing harmonies to show {#sec:chooseharmonies}

Even after screening harmonies from the list of all possible combinations of cyclic time granularities that could be plotted together, the number of combinations quickly becomes too large and thus, it is beneficial useful to rank these harmonies. Ranking could be based on how well a harmony pair displays the variation in distributions of the measured variable within and across each levels of the cyclic granularities. For each facet level, the most interesting display is the one in which the difference in the distribution of the measured variable between the x-axis categories is maximum. The harmony pair which has more variation between x-axis levels for all facet levels, on average, is likely to be a good measure of how important a harmony pair is displaying variation in the distributions. Probability distributions are represented through sample percentiles instead of kernel density estimate so that there is minimal dependency on selecting kernel or bandwidth. Distances between these probability distributions of the measured variable across x-axis levels are computed using Jenson-Shannon distances which is a symmetric divergence obtained from Kullback-Leibler divergence measure.

@@

Restructuring time from linear to cyclic time granularities leads to re-organisation of the data structure, where each level of a cyclic granularity corresponds to multiple values of the measured variable. This is unlike the case of linear granularities where each time point corresponds to an unique value of the measured variable. It is common to see summarization of these multiple values through aggregation or an unique summary statistic like mean or median to have each level of cyclic granularity correspond to an unique value of the measured variable. However, this approach hides the distributions of the measured variable induced by the re-organised data structure. Summarizing the distribution of these multiple observations across cyclic granularities using some summary statistics, each of which can highlight different features of the data or through probability densities is a valid way to compare and contrast periodicities in the data.<!--Any attempt to encode all or many cyclic granularities together to develop insights on periodicity might become clumsy. Instead, the big problem is broken down by focusing on two cyclic granularities at a time Summarizing properties of the distribution of the measured variable can give valuable indication of features.-->

@@

- Data: <$C_1$, $C_2$, $v$>
- Aesthetic mapping (mapping of variables to elements of the plot): $C_1$ mapped to $x$ position and $v$ to $y$ position
- Statistical transformation (data summarization): any descriptive or smoothing statistics that summarizes distribution of $v$
- Geometric objects (physical representation of the data): any geometry displaying distribution, for example, boxplot, letter value, violin, ridge or highest density region plots
- Facet (split plots):  $C_2$

@@

Choice of plots are dictated by the statistical transformations and geometric objects used for the visualization. The basic plot choice for our data structure is the one that can display distributions<!--In Section \ref{sec:measured-variable}, it is discussed that these densities could be estimated.-->. We will discuss few conventional and recent ways to plot distributions using both Kernel density estimates and descriptive statistics. Descriptive statistics based displays include box plots [@Tukey1977-jx] or different variations of it like notched box plots [@Mcgill1978-hg]. More recent ways are the letter-value box plot [@Hofmann2017-sg] or quantile plots which display quantiles instead of quartiles in a traditional boxplot. Kernel density based plots for displaying a distribution include violin plots [@Hintze1998-zi], summary plot [@Potter2010-qc], ridge line plots, highest density regions (HDR) box [@Hyndman1996-ft]. Each type of density display has different parameters, that need to be estimated given the data. Each is equipped with some benefits and challenges which should be borne in mind while using them for exploration.

@@

If there are levels of cyclic granularities plotted across the  x-axis which are not spanned by levels of cyclic granularities plotted across facets or vice versa, we will have empty sets leading to potential ineffective graphs. We hypothesize that the synergy of these cyclic granularities are thus playing a role while deciding if the resulting plot would be a good candidate for exploratory analysis. Harmonies are pairs of granularities that do not contain empty combinations and thus aid exploratory data analysis, and clashes are pairs that do not aid the process of exploratory analysis because they contain empty combinations. As a result, we should avoid plotting clashes. For illustration, \autoref{fig:allFig} (a) shows the distribution of half-hourly electricity consumption through letter value plot across months of the year faceted by quarter-of-year. This plot does not work because every quarter does not correspond to all months of the year, for example, the first quarter would not correspond to the month December of an year.

@@

```{r allFig, out.width="100%", fig.align= 'left',echo=FALSE, eval=TRUE, warning=FALSE,message = FALSE, fig.cap="Distribution of energy consumption displayed through letter value plots. Plot (a) displays consumption across month-of-year faceted by quarter-of-year, (b) across weekday/weekend faceted by quarter-of-year and (c) across quarter-of-year faceted by weekday/weekend. Plots (b) and (c) both show harmonies since each quarter consists of both weekdays and weekends. Conversely, weekend and weekday can occur in every quarter. Analysts should avoid plotting such clashes. Plot (b) helps to compare weekends and weekdays for each quarter. It can be seen that for every quarter weekend and weekday consumption are fairly similar except for the second quarter where the letter values below D and E behave differently, whereas, Plot (c) helps to compare quarters within weekdays and weekends. For example, the quartile spread of consumption shrinks/lowers from first to fourth quarter for weekdays, whereas this pattern is not true for weekends. Plot (a) shows a clash since all quarters do not correspond to all months of the year. ", fig.show = 'hold'}
